---
title: "Monthly AAVSO groups with Hurdle model"
author: "Mark Heiple..."
date: "`r format(Sys.time(), "%d %B, %Y")`"
output:
  html_document:
    df_print: paged
keep_tex: yes
number_sections: yes
highlight: pygments
fig_caption: no
toc: yes
toc_depth: 4
---

```{r, include=FALSE, purl=FALSE}
library(knitr)
source("app_functions.R")

#set echo=FALSE to suppress code chunks appearing in text
#setting it in the individual code chunks below doesn't appear to work
knitr::opts_chunk$set(echo = FALSE)

#code chunks are defined here
#knitr::read_chunk('report_test.R')

```


```{r, echo=FALSE,include=FALSE, cache=FALSE}
da1=read.csv("aavso_month_.csv",header=TRUE)
head(da1)

#as strings - use this
ds = sapply(seq(nrow(da1)), FUN=function(i) {
  s = sprintf("%d-%.02d-%.02d",da1$year[i],da1$month[i],da1$day[i])
})

#get periods
ppdf = findPeriod(da1$g)

#use this period
freq = ppdf$period[2]

#create new data frame using the dates as strings
da2 = data.frame(date=ds,g=ts(da1$g, frequency=freq),s=ts(da1$s, frequency=freq),w=ts(da1$W, frequency=freq))
```

```{r, echo=FALSE, include=FALSE}

arima_ft = function(y, d=NA, KK=2, df_post_probs=NULL) {

  #season
  periods = attr(y,"msts")
  if( is.null(periods) ) {
    periods = frequency(y)
  }
  nPeriods = length(periods)

  fuur = fourier(y, K = rep(KK,nPeriods))
  
  if( !is.null(df_post_probs) ) {
    xreg = as.matrix(cbind(fuur,df_post_probs))
  } else {
    xreg = fuur
  }
  m = auto.arima(y, seasonal=FALSE, d, xreg=xreg, stepwise = TRUE, max.p=10,max.q=10,max.order=20)

  return(m)  
}

#function for doing EDA
eda = function(y, dates, ts_name) {

  #plot raw ts
  cat("\n\n##Time Series\n\n")
  plot_ts(y, tdx=dates, main=ts_name)
  
  #plot seasonal decomposition
  cat("\n\n##Seasonal Decomposition\n\n")
  dd = stl(y, s.window="periodic")
  plot(dd, main = sprintf("%s, season = %d", ts_name, freq))
  
  ## ACF/PACF:
  cat("\n\n##ACF/PACF\n\n")
  plot_acf_pacf(y,main=ts_name)
  cat('\n\n')
  
  ## Distributions
  cat("\n\n##Distributions\n\n")
  plot_dist(y)
  cat('\n\n')
  
  ## Histogram
  cat("\n\n##Histogram vs Various Distributions\n\n")
  cat(sprintf("\n\nNote: Variance = %.2f, which is significantly greater than the mean (%.2f), which indicates that the data is closer to a negative binomial distribution.\n\n",round(var(y),digits=2),round(mean(y),digits=2)))
  if(sum(which(y<0))==0) {
    plot_hist(y)
  } else {
    plot_hist_diff(y)
  }
  cat('\n\n')

  #summary   
  cat("\n\n##Summary Statistics\n\n")
  df = summary_table(y,ts_name)
  print(knitr::kable(df,digits=3,caption=sprintf("%s: Summary Statistics", ts_name)))
  cat('\n\n')
  
  #hypothesis tests
  cat(sprintf("\n\n##Hypothesis Tests\n\n", ts_name))
  df = tests(y)
  print(knitr::kable(df,digits=3,caption=sprintf("%s: Hypothesis Tests",ts_name)))
  cat('\n\n')
}

nseasons = 10
freqs = head(ppdf,nseasons)$period
y = msts(da2$g, seasonal.periods = freqs)

#experiment with arima model to try deteriming # of significant lags and seasons
#this takes a long time
#m = arima_ft(y, d=1, KK=2)
```
\newpage
#Determine # of parameters for Hurdle Model
I did some experimenting with the ARIMA model to determine how many parameters to use in a hurdle model. I found that more than 4 seasonal periods and more then 1 harmonic tested as insignificant. However, the tests assume that the residuals are iid normal, which they are not. It also showed an ARMA order greater than (3,3) was not significant. auto.arima tested models up to (10,10).

The hurdle model uses the following `r nseasons` seasonal periods:
`r knitr::kable(head(ppdf,nseasons))`

```{r,echo=FALSE, include=TRUE, results='asis'}
#summary(m)
#parameterTest(m,da2$w)
#eda(residuals(m),ds,ts_name="ARIMA residuals")
```

\newpage
#Hurdle Model

```{r,echo=FALSE, include=TRUE}

#build a data frame to include the seasonal periods + lagged values of the counts
make_lags = function(x, nlags, KK=1) {
  
  n = length(x)
  
  #build lags
  dflag = lapply(seq(nlags,by=-1), function(i){
    #return(tail(da2$w,i))
    df = as.data.frame(head(tail(x,n-i+1),n-nlags))
    names(df) = sprintf("Lag-%d",nlags-i+1)
    return(df)
  })
  
  #seasonal components
  periods = attr(x,"msts")
  if( is.null(periods) ) {
    periods = frequency(x)
  }
  nPeriods = length(periods)
  fuur = as.data.frame(tail(fourier(x, K = rep(KK,nPeriods)),n-nlags))

  y = data.frame(tail(x,n-nlags))
  names(y)="y"
  
  df = data.frame(c(y,dflag,fuur))
  
  return(df)
}

make_forecast <- function(m) {
  f = list(
    x = df$y,
    fitted = msts(m$fitted.values, seasonal.periods = freqs)
  )

  return(f)  
}

nlags = 10
df = make_lags(y, nlags)
#head(df)

```

Even though the ARIMA model indicated that more than 4 seasons was not significant, this did not carry over the hurdle model. More seasons were much more significant to the count model than the zero model.

More lags do not seem to help at all. I tested 3, as indicated by the ARIMA model, 10, 100, and 1000. 10 and 100 showed no improvement, 1000 was significantly worse. With a large number of lags, the fitted values no longer match the seasonality of the original data.

Adding seasonal periods has a much greater impact on the model than adding lags.

Fitted a model with `r nlags` lags and `r nseasons` seasons.

I attempted to fit a hurdle model to W monthly data, but it failed with the following error:

Error in solve.default(as.matrix(fit_zero$hessian)) : 
  Lapack routine dgesv: system is exactly singular: U[1,1] = 0


```{r,echo=FALSE, include=TRUE}
#build the hurdle model

library(pscl)
m <- hurdle(y ~ ., data = df, dist = "negbin", zero.dist = "binomial")

# same as this:
# m <- hurdle(visits ~ ., data = nmes, dist = "poisson", zero.dist = "binomial")
summary(m)
AIC(m)
BIC(m)


#how many zeros are predicted = 3040
sum(predict(m, type = "prob")[,1])

#actual = 3040
sum(df$y < 1)  
```

###Model Fit

Plots of the fitted data and original data, and the Rootogram showing goodness of fit. It shows that it under predicts small count values and over predicts larger counts.

```{r,echo=FALSE, include=TRUE, results='asis'}
library(countreg)

#AIC = `r AIC(m)`
#BIC = `r BIC(m)`


f = make_forecast(m)

cat("\n\n")
plot_fit(f,ds,main="Hurdle")
cat("\n\n")
plot_fit(f,ds,first=1000,main="G Monthly, First 1000")
cat("\n\n")
plot_fit(f,ds,last=400,main="G Monthly, Last 400")
cat("\n\n")
plot_fit(f,ds,last=144,main="G Monthly, Last 144")

cat("\n\n")
rootogram(m, max = max(df$y))
#cat("\n\n")
#rootogram(m, max = mean(df$y))


```

###Residuals
Residuals look a little closer to normal than the previous time series models. ACF still shows an unresolved seasonal period.

```{r,echo=FALSE, include=TRUE, results='asis'}
cat("\n\n")
plot_dist(residuals(m))
cat("\n\n")
acf(residuals(m),max(freqs))
cat("\n\n")
#tsdiag(m)
#cat("\n\n")
```

